\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb}

\title{Knowledge Representation and Reasoning \\ Project 4: Actions with Cost \\ Checkpoint 2}
\author{Jakub Seliga, Michał Szewczak, Sebastian Trojan, Wiktoria Koniecko}
\date{}

\newcommand{\fluents}{\mathcal{F}}
\newcommand{\actions}{\mathcal{A}c}
\newcommand{\states}{\Sigma}
\newcommand{\parr}{\par\vspace{5mm}\noindent}

\begin{document}

\maketitle

\section*{Introduction}

The field of Knowledge Representation and Reasoning (KRR) focuses on formalizing information so that a computer system can utilize it to solve complex tasks, such as simulating the evolution of a dynamic environment. Within this domain, action languages provide a structured way to model how autonomous agents interact with their surroundings through sequences of operations. By defining the logic behind state transitions, these systems enable automated reasoning about the consequences, feasibility, and resource requirements of specific plans.
The primary objective of this project is to formalize and implement a logic-based framework for representing and reasoning about dynamic systems within the $\mathbb{DS}_{4}$ class. The class is an extension of class of dynamic systems described by Action Language $\mathbb{A}$,  and has following assumptions:
\begin{enumerate}
    \item \textbf{Inertia law:} Fluents persist in their state unless changed by an action.
    \item \textbf{Complete information:} The system assumes full knowledge of all actions and all fluents.
    \item \textbf{Branching model of time:} The temporal structure allows for multiple possible future paths.
    \item \textbf{Deterministic and sequential actions:} Only one action occurs at a time, and every action has a single, predictable outcome.
    \item \textbf{Characteristics of actions:}
    \begin{itemize}
        \item \textit{Precondition:} A set of literals representing the conditions for execution; if not met, the action results in an empty effect.
        \item \textit{Postcondition:} The specific effect of an action, represented as a literal.
        \item \textit{Cost:} A cost $\kappa \ge 1$ is required to perform an action. However, $\kappa = 0$ if the cost is unspecified or if the action results in an empty effect.
    \end{itemize}
    \item \textbf{Universal performance:} In every state, all actions are performed.
    \item \textbf{Partial descriptions:} The system allows for incomplete descriptions of any given state.
    \item \textbf{No constraints:} No state or domain constraints are admitted in this class.
\end{enumerate}
The difference from standard Action Language $\mathbb{A}$ is the cost associated with an action.
\section{Action language syntax}
Here \(\actions\) is the set of all actions and \(\fluents\) is the set of all fluents.

\subsection{Value statement}
\[\bar{f} ~\text{after}~ A_1, \ldots, A_n\]
Where \(\bar{f}\in \fluents\) is a fluent and \(A_1, \ldots, A_n \in \actions\) are actions.\\
Abbreviation when \(n=0\): initially \(\bar{f}\)

\subsection{Effect statement}
\[A ~\text{causes}~ \bar{f} ~\text{if}~ g_1 ,\ldots, g_k\]
Where \(\bar{f},g_1,\ldots,g_k \in \fluents\) are fluents and \(A\in \actions\) is an action.\\
Abbreviation when \(k=0\): \(A_1\) causes \(\bar{f}\)

\subsection{Cost statement}
\[A ~\text{costs}~ \kappa\]
Where \(A\in \actions\) is an action and \(\kappa>0\).

\section{Action language semantics}
\noindent A \textbf{signature} \(\Upsilon\) consists of:
\begin{itemize}
    \item A finite set of fluents $\fluents$
    \item A finite set of actions $\actions$
\end{itemize}

\parr
A non-empty set \(D\) of value/effect/cost statements is called an \textbf{action domain}.
Let \(D_{trans} \subseteq D\) be the set of all value and effect statements in \(D\), and
\(D_{cost} \subseteq D\) be the set of all cost statements in \(D\).


Model construction is based solely on \(D_{trans}\).
Cost statements \(D_{cost}\) play no role in determining whether a structure is a model;
they are used only in a second phase, after the transition function \(\Psi\) has been
fully determined, to label transitions with their costs.

\parr
A \textbf{transition function} is a mapping \(\Psi: \actions\times\states\rightarrow\states\).
The value \(\Psi(A, \sigma)\) is the state resulting from performing action \(A\) in state \(\sigma\).
In case of the \(\mathbb{DS}4\) class of dynamic systems the mapping \(\Psi\) must be defined for all
actions and states because performing an action is always possible.

\parr
Let \(\mathcal{L}\) be an action language for a \(\mathbb{DS}4\) dynamic system.
A \textbf{structure for} \(\mathcal{L}\) is a pair \(S=(\Psi, \sigma_0)\), where:
\begin{itemize}
    \item \(\Psi\) is a transition function
    \item \(\sigma_0\in\states\) is the initial state
\end{itemize}

\parr
A statement \(s \in D_{trans}\) \textbf{is true in structure} \(S\) (in symbols \(S\models s\)) iff:
\begin{itemize}
    \item \(s\) is of the form \[\bar{f} ~\text{after}~ A_1, \ldots, A_n\]
    and \(\Psi^\star((A_1, \ldots, A_n),\sigma_0)\models \bar{f}\)
    \item \(s\) is of the form \[A ~\text{causes}~ \bar{f} ~\text{if}~ g_1 ,\ldots, g_k\]
    and for every \(\sigma\in\states\) such that \(\sigma\models g_i\) for all \(i\):
    \(\Psi(A, \sigma)\models\bar{f}\)
\end{itemize}

\parr
A structure \(S=(\Psi, \sigma_0)\) is a \textbf{model} of action domain \(D\) in a language over
the signature \(\Upsilon=(\fluents,\actions)\) iff:
\begin{itemize}
    \item \((\forall s\in D_{trans})~ S\models s\)
    \item for every effect statement \(s\) of the form
    \[A ~\text{causes}~ \bar{f} ~\text{if}~ g_1 ,\ldots, g_k\]
    and for every \(\sigma\in\states\), if one of the following holds:
    \begin{itemize}
        \item \(s\in D_{trans}\) and \((\exists i\in\{1,\ldots,k\})~ \sigma\nvDash g_i\)
        \item \(s\notin D_{trans}\)
    \end{itemize}
    then \(\sigma \models \bar{f}\) iff \(\Psi(A,\sigma)\models \bar{f}\)
\end{itemize}

\parr
Let \(S = (\Psi, \sigma_0)\) be a model of \(D\) (determined using \(D_{trans}\) only).
The \textbf{cost function}
\(Cost: \actions\times\states\rightarrow\mathbb{N}\) is then uniquely determined by \(\Psi\)
and \(D_{cost}\) as follows:
\[Cost(A, \sigma) = \begin{cases}
\kappa & \text{if } (A ~\text{costs}~ \kappa) \in D_{cost}
        ~\text{and}~ \Psi(A, \sigma)\neq\sigma \\
0      & \text{otherwise}
\end{cases}\]
The \emph{otherwise} branch covers two cases mandated by assumption A5:
(a) no cost statement for \(A\) appears in \(D_{cost}\), and
(b) \(A\) has an empty effect in state \(\sigma\) (i.e.\ \(\Psi(A,\sigma)=\sigma\)).
In both cases \(Cost(A,\sigma)=0\).

Since \(Cost\) is uniquely determined by \(\Psi\) and \(D_{cost}\), we write the full structure
as a triple \(S=(\Psi, \sigma_0, Cost)\) as a notational convenience, understanding \(Cost\)
as a derived component.

\parr
A \textbf{program} \(P\) is a sequence of actions to be performed starting from the initial state.
The cost of a program \(P=(A_1,\ldots,A_k)\) is defined as:
\[Cost^\star((A_1,\ldots,A_k)) = \sum_{i=1}^k Cost\!\left(A_i,\, \Psi^\star((A_1,\ldots,A_{i-1}), \sigma_0)\right)\]
where \(\Psi^\star((), \sigma_0) = \sigma_0\) by convention.

\section{Query Language statements}

\subsection*{Q1: Goal satisfaction}
\[\gamma ~\text{after}~ P\]
Where \(\gamma\) is the goal condition (a set of literals) and \(P\) is a program.

\subsection*{Q2: Executability with Cost}
\[P ~\text{executable with cost}~ \kappa\]
Where \(\kappa > 0\) and \(P\) is a program.

\subsection*{Q3: Executability with exact Cost}
\[P ~\text{executable with exact cost}~ \kappa\]
Where \(\kappa > 0\) and \(P\) is a program.

\section{Semantics of the query language}

\subsection{Partial Initial States}

A partial initial state represents a set of complete states (completions), each assigning a truth value to every fluent while remaining consistent with the given literals.
\\Execution of a program $P$ from a partial state $\sigma_0$ induces a set of possible models:
\[
S_i=(\Psi^i, \sigma_0^i, Cost)
\]

\subsection*{Q1: Goal satisfaction}
A query of the form:
\[\gamma ~\text{after}~ P\]
is a consequence of action domain \(D\) iff for every model \(S_i=(\Psi^i, \sigma_0^i, Cost)\):
\[
(\forall \bar{f}\in\gamma)~~ \Psi^i(P, \sigma_0^i) \models \bar{f}
\]

\subsection*{Q2: Executability with Cost}
A query of the form:
\[P ~\text{executable with cost}~ \kappa\]
is a consequence of action domain \(D\) iff for every model \(S_i=(\Psi^i, \sigma_0^i, Cost)\):
\[
Cost(P, \sigma_0^i) \leq \kappa
\]

\subsection*{Q3: Executability with exact Cost}
A query of the form:
\[P ~\text{executable with exact cost}~ \kappa\]
is a consequence of action domain \(D\) iff for every model \(S_i=(\Psi^i, \sigma_0^i, Cost)\):
\[
Cost(P, \sigma_0^i) = \kappa
\]

\section*{Examples}

\subsection*{Example 1}
Let's consider the following statements:
\[
\text{initially}~ \neg doorOpen
\]
\[
openDoor ~\text{causes}~ doorOpen ~\text{if}~ hasKey
\]
\[
openDoor ~\text{costs}~ 5
\]

\subsubsection*{Program}
\[
P = (openDoor)
\]

\subsubsection*{Completions of $\sigma_0$}
\[
\sigma_0^1 = \{\neg doorOpen, hasKey\}
\]
\[
\sigma_0^2 = \{\neg doorOpen, \neg hasKey\}
\]

\subsubsection*{Execution}

\paragraph{Case 1: $\sigma_0^1$}

Precondition holds, so:
\[
\sigma_1^1 = \{doorOpen, hasKey\}
\]

Cost:
\[
Cost^\star(P) = 5
\]

\paragraph{Case 2: $\sigma_0^2$}

Precondition does not hold, so:
\[
\sigma_1^2 = \{\neg doorOpen, \neg hasKey\}
\]

Cost:
\[
Cost^\star(P) = 0
\]

\subsubsection*{Final States}
\[
\{\sigma_1^1, \sigma_1^2\}
\]

\subsubsection*{Query Results}

\paragraph{Q1: Goal Satisfaction}
\[
\{doorOpen\} ~\text{after}~ P
\]
is false because $doorOpen$ does not hold in all final states.

\paragraph{Q2 and Q3: Executability with Cost}

\[
P ~\text{executable with exact cost}~ 5
\]
is false since $Cost^\star(P)$ is not always 5.

\[
P ~\text{executable with cost}~ 5
\]
is true since for all executions:
\[
Cost^\star(P) \leq 5
\]

\[
P ~\text{executable with cost}~ 0
\]
is false since there exists an execution with cost 5.


\subsection*{Example 2}
Let's consider the following statements:
\[
\text{initially}~ \neg doorOpen
\]
\[
\text{initially}~ hasKey
\]
\[
openDoor ~\text{causes}~ doorOpen ~\text{if}~ hasKey
\]
\[
openDoor ~\text{costs}~ 5
\]

\subsubsection*{Program}
\[
P = (openDoor, openDoor)
\]

\subsubsection*{Execution}

Initial state:
\[
\sigma_0 = \{\neg doorOpen, hasKey\}
\]

\paragraph{Action 1}

Precondition holds, so:
\[
\sigma_1 = \{doorOpen, hasKey\}
\]

Cost:
\[
Cost(openDoor, \sigma_0) = 5
\]

\paragraph{Action 2}

Precondition holds, but no change occurs:
\[
\sigma_2 = \{doorOpen, hasKey\}
\]

Cost:
\[
Cost(openDoor, \sigma_1) = 0
\]

\paragraph{Total Cost}
\[
Cost^\star(P) = 5
\]

\subsubsection*{Query Results}

\paragraph{Q2: Executability with Cost}
\[
P ~\text{executable with cost}~ 5
\]
is true.

\paragraph{Q3: Executability with exact Cost}
\[
P ~\text{executable with exact cost}~ 10
\]
is false since:
\[
Cost^\star(P) = 5 \neq 10.
\]

\subsection*{Example 3}
Let's consider the following statements:
\[
\text{initially}~ \neg doorOpen
\]
\[
doorOpen ~\text{after}~ openDoor
\]
\[
openDoor ~\text{causes}~ doorOpen ~\text{if}~ hasKey
\]
\[
openDoor ~\text{costs}~ 5
\]

\subsubsection*{Program}
\[
P = (openDoor)
\]

\subsubsection*{Completions of $\sigma_0$}
\[
\sigma_0^1 = \{\neg doorOpen, hasKey\}
\]
\[
\sigma_0^2 = \{\neg doorOpen, \neg hasKey\}
\]

However, $\sigma_0^2$ is not consistent with the action domain.

Indeed, since $hasKey$ does not hold, the effect does not apply, and by inertia:
\[
\Psi(openDoor, \sigma_0^2) \models \neg doorOpen
\]

But the value statement requires:
\[
\Psi(openDoor, \sigma_0^2) \models doorOpen
\]

This is a contradiction. Therefore, $\sigma_0^2$ is not allowed.

\subsubsection*{Execution}

Thus, the only possible initial state is:
\[
\sigma_0 = \{\neg doorOpen, hasKey\}
\]

Precondition holds, so:
\[
\sigma_1 = \{doorOpen, hasKey\}
\]

Cost:
\[
Cost^\star(P) = 5
\]

\subsubsection*{Final States}
\[
\{\sigma_1\}
\]

\subsubsection*{Query Results}

\paragraph{Q1: Goal Satisfaction}
\[
\{doorOpen\} ~\text{after}~ P
\]
is true.

\paragraph{Q2: Executability with Cost}
\[
P ~\text{executable with cost}~ 5
\]
is true.

\paragraph{Q3: Executability with exact Cost}
\[
P ~\text{executable with exact cost}~ 5
\]
is true.

\subsection*{Example 4}
Let's consider the following statements:
\[
\text{initially}~ \neg doorOpen
\]
\[
\text{initially}~ \neg hasKey
\]
\[
doorOpen ~\text{after}~ openDoor
\]
\[
openDoor ~\text{causes}~ doorOpen ~\text{if}~ hasKey
\]
\[
openDoor ~\text{costs}~ 5
\]

\subsubsection*{Program}
\[
P = (openDoor)
\]

\subsubsection*{Execution}

There is no model satisfying all statements, because:
\begin{itemize}
\item $\sigma_0 \models \neg hasKey$
\item the effect requires $hasKey$
\item the value statement enforces $doorOpen$ after $openDoor$
\end{itemize}

These constraints are inconsistent.

\subsubsection*{Query Results}

All queries are undefined since no model exists.




\end{document}